---
title: "METAMATCH-Sleep: methods walkthrough"
subtitle: "Leveraging NLP to Identify Domain-Specific Variables in Large-Scale Cohort Metadata"
author: "Ben J Draper"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 6, dpi = 120
)
```

This document walks through the analysis end to end and knits to a single HTML
page. It is a narrative companion to the scripts in `analysis/`, not a
replacement for them — each section points at the script that does the work, and
that script is where the detailed rationale lives.

Run it with `rmarkdown::render("methods_walkthrough.Rmd")`.

```{r libraries}
source(here::here("R", "setup.R"))
```

---

# 1. The problem

Cohort studies describe the same construct in wildly different words. Finding
every sleep variable across 20 studies means reading 86,681 variable
descriptions, which nobody does by hand.

The difficulty is not just volume, it is that volume is unevenly distributed and
unrelated to study size.

```{r cohort-landscape}
cohorts <- read_csv(file.path(PATH_RAW, "cohort_summary.csv"), show_col_types = FALSE) %>%
  mutate(cohort = recode_cohorts(cohort), Estimated_N = as.numeric(Estimated_N)) %>%
  arrange(desc(n_vars))

cohorts %>%
  summarise(
    cohorts       = n(),
    variables     = sum(n_vars),
    participants  = sum(Estimated_N),
    median_vars   = median(n_vars),
    smallest      = min(n_vars),
    largest       = max(n_vars)
  ) %>%
  knitr::kable(format.args = list(big.mark = ","))
```

Variable counts span three orders of magnitude — CamPaIGN documents 32, ELSA
documents 44,826 — and the correlation with population size is weak. Metadata
volume reflects documentation practice, not study size, so sleep variables
cannot be found by looking only at the large cohorts.

**Figure 1** — `analysis/02_fig1_cohort_landscape.R`

---

# 2. Retrieval: the SLEEPTALKING corpus and TF-IDF

A reference dictionary of 2,032 sleep terms was pulled from the UMLS
Metathesaurus in June 2022, starting from CUI `C0037313` ("Sleep").

```{r corpus}
umls <- read_csv(file.path(PATH_REFERENCE, "sleeptalking_corpus.csv"), show_col_types = FALSE)
tibble(terms = nrow(umls), unique_names = n_distinct(umls$name)) %>% knitr::kable()
head(umls$name, 8)
```

Every variable description was scored against every UMLS term using character
n-gram TF-IDF and cosine similarity, keeping the best match per description.

Why TF-IDF rather than exact matching? Because exact matching finds almost
nothing:

```{r why-not-exact}
sleep_dic <- read_dpuk_csv(file.path(PATH_INTERIM, "sleep_dic_man.csv"))

sleep_dic %>%
  filter(Similarity < 0.999) %>%
  arrange(desc(Similarity)) %>%
  select(`DPUK description` = From, `UMLS term` = To, Similarity) %>%
  slice(c(1:3, 200:202, 500:502)) %>%
  knitr::kable(digits = 3)
```

These pairs share meaning while sharing very little literal text. TF-IDF over
n-grams tolerates the abbreviation and punctuation differences; exact matching
would reject all of them.

**Threshold.** Matches at cosine ≥ 0.5 were kept and then read individually.
The cut is deliberately permissive, favouring recall and letting the expert
reject false positives.

```{r similarity-summary}
sleep_dic %>%
  summarise(matches = n(), min = min(Similarity), mean = mean(Similarity), max = max(Similarity)) %>%
  knitr::kable(digits = 3)
```

**Figure 2** — `analysis/01_tfidf_umls_match.R`, `analysis/03_fig2_similarity_duplication.R`

## 2a. How good were the matches?

This is the number the paper's conclusion rests on.

```{r umls-qc}
qc <- read_excel(file.path(PATH_CURATED, "sleep_vars_umls_qc_manual.xlsx")) %>%
  filter(bjd_UMLS_QC %in% c("YES", "NO"))

qc %>%
  count(`UMLS assignment correct?` = bjd_UMLS_QC) %>%
  mutate(`%` = round(100 * n / sum(n))) %>%
  knitr::kable()
```

**Fewer than half.** A high cosine score is not the same as a correct match, and
no threshold fixes that — it is a limitation of lexical similarity, not of the
cut-off. This is the evidence for keeping an expert in the pipeline.

---

# 3. Semantic representation: BERT [CLS] embeddings

TF-IDF matches shared vocabulary; it cannot tell that *"trouble getting off to
sleep"* and *"difficulty initiating sleep"* mean the same thing. Each cleaned
description was therefore embedded with `bert-base-uncased`, taking the `[CLS]`
token as a whole-sentence summary.

> **Note.** The preprint states MiniLM-L6-v2. The code uses `bert-base-uncased`,
> and the 768-dimensional embedding files confirm it. See `REPRODUCIBILITY.md`.

```{r embeddings}
sleep_emb <- read_cls_embeddings(file.path(PATH_INTERIM, "cls_embeddings_sleep754.csv"))
assert_aligned(sleep_emb, sleep_dic, "sleep dictionary")

tibble(
  rows       = nrow(sleep_emb),
  dimensions = ncol(sleep_emb) - 1
) %>% knitr::kable()
```

Cleaning collapses many descriptions onto identical strings, which produce
bit-identical embeddings. Deduplicating on the embedding vector — the
representation the clustering actually sees — gives the 337 figure quoted
throughout the paper.

```{r dedup}
sleep_unique <- distinct_embeddings(sleep_emb)
```

---

# 4. Projection: what structures the space?

```{r tsne, fig.height = 5}
set.seed(SEED_TSNE_LOCAL)
tsne <- Rtsne(embedding_matrix(sleep_unique), dims = 2, perplexity = 30,
              max_iter = 1000, verbose = FALSE)

PROMINENT <- c("CaPS", "CFAS", "CamCAN", "ELSA", "Whitehall", "AMPLE", "MRC")

sleep_dic$session_id <- as.character(seq_len(nrow(sleep_dic)))

layout <- tibble(tsne_1 = tsne$Y[, 1], tsne_2 = tsne$Y[, 2],
                 session_id = sleep_unique$session_id) %>%
  left_join(sleep_dic, by = "session_id") %>%
  mutate(cohort = recode_cohorts(cohort),
         cohort = if_else(cohort %in% PROMINENT, cohort, "other"),
         n_words = str_count(From, "\\S+"))

ggplot(layout, aes(tsne_1, tsne_2, colour = cohort)) +
  geom_point(size = 2) +
  scale_colour_manual(values = COHORT_COLOURS) +
  labs(title = "Sleep definitions coloured by cohort",
       x = "Dimension 1", y = "Dimension 2") +
  theme_minimal()
```

The paper's main negative result is visible here: definitions group by **cohort**
rather than by sleep construct. Cohort house style dominates semantic content.

The second driver is description length — an artefact of using untuned `[CLS]`
vectors, whose magnitude tracks token count:

```{r length-effect, fig.height = 4.5}
ggplot(layout, aes(tsne_1, tsne_2, colour = n_words)) +
  geom_point(size = 2) +
  scale_colour_viridis_c(name = "Words in\ndescription", option = "magma", direction = -1) +
  labs(title = "The same projection, coloured by description length",
       x = "Dimension 1", y = "Dimension 2") +
  theme_minimal()
```

**Figure 3** — `analysis/04_fig3_tsne.R`

---

# 5. Clustering vs expert consensus

The unsupervised route (UMAP → HDBSCAN) and the expert route (keyword seeding →
consensus review) were run on the same 374 full-sentence definitions so they
could be compared.

```{r categories}
counts <- read_csv(file.path(PATH_CURATED, "consensus_category_count.csv"), show_col_types = FALSE)

counts %>% arrange(desc(Count)) %>% knitr::kable()
```

```{r category-plot, fig.height = 5}
category_colours <- c(rep("#440099", 4), UOS_SEQUENCE)
names(category_colours) <- counts %>% arrange(Count) %>% pull(Category)

ggplot(counts, aes(Count, reorder(Category, Count), fill = Category)) +
  geom_col() +
  scale_fill_manual(values = category_colours) +
  labs(title = "Expert consensus categories",
       x = "Number of sleep variables", y = NULL) +
  theme_minimal() +
  theme(legend.position = "none")
```

The three commonest categories — Sleep Difficulty, Sleep Latency & Waking, Sleep
Duration — are the constructs a harmonisation effort would target first.

The unsupervised clusters are coarser and partly artefactual by comparison: one
is HDBSCAN's noise group (proxy-respondent and undefined items) and another is
defined largely by description length. Two caveats on that panel are important
enough to repeat here:

- Clustering runs on the **category-expanded** frame, so multi-label definitions
  are weighted more heavily. Panel 4a is not purely unsupervised.
- Its UMAP layout depends on RNG state left by a preceding t-SNE call, so exact
  cluster sizes are not reproducible across package versions.

Both are documented in `REPRODUCIBILITY.md`; `analysis/05_fig4_clusters_categories.R`
detects when it has not reproduced the published clustering and withholds the
expert cluster names rather than applying them to a different result.

**Figure 4** — `analysis/05_fig4_clusters_categories.R`

---

# 6. Conclusion

NLP reduced 86,681 variable descriptions to 337 semantically unique sleep
variables — a search a researcher can actually complete. But fewer than half the
automated UMLS assignments were correct, the embedding space is organised by
cohort and sentence length rather than by sleep construct, and the unsupervised
clusters needed human interpretation to mean anything.

The method is a powerful filter, not a replacement for domain expertise. Where
each stage could be strengthened for a version 2 is set out in
[`docs/methods_v2_2026.html`](docs/methods_v2_2026.html).

```{r session-info}
sessionInfo()
```
